---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# `dearseq` <a><img src='man/figures/logo.svg' align="right" height="139" /></a>

[![BioC status](https://www.bioconductor.org/shields/build/release/bioc/dearseq.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/dearseq/)
[![BioC dev status](https://www.bioconductor.org/shields/build/devel/bioc/dearseq.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/dearseq/)
[![R build status](https://github.com/borishejblum/dearseq/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/borishejblum/dearseq/actions)
[![Codecov test coverage](https://codecov.io/gh/borishejblum/dearseq/branch/devel/graph/badge.svg)](https://app.codecov.io/gh/borishejblum/dearseq?branch=devel)
[![Download](https://www.bioconductor.org/shields/downloads/release/dearseq.svg)](https://bioconductor.org/packages/stats/bioc/dearseq)
<!-- [![Downloads](https://cranlogs.r-pkg.org/badges/dearseq?color=blue)](https://www.r-pkg.org/pkg/dearseq) -->

**Authors:** [Boris Hejblum](https://borishejblum.science), Denis Agniel, Marine Gauthier, Mélanie Huchon, Iban Larrouture<br/>
**License:** [GNU GPL-2](https://opensource.org/licenses/gpl-2-0)


## Overview

`dearseq` is a package for analyzing RNA-seq data. The 2 main functions of the package are `dear_seq()` and `dgsa_seq()`: 

* **Gene-wise Differential Analysis of RNA-seq data** can be performed using the function `dear_seq()`. 
* **Gene Set Analysis of RNA-seq data** can be performed using the function `dgsa_seq()`. 

The methods implemented in this package are detailed in the following articles:

> Gauthier M, Agniel D, Thiébaut R & Hejblum BP (2020). dearseq: a variance component score test for RNA-Seq differential analysis that effectively controls the false discovery rate, *NAR Genomics and Bioinformatics* **2**(*4*):lqaa093 [DOI: 10.1093/nargab/lqaa093](https://doi.org/10.1093/nargab/lqaa093) [bioRxiv:635714](https://www.biorxiv.org/content/10.1101/635714v1)

`r emoji::emoji("recycling_symbol")` Reproducible code files for the results presented in Gauthier *et al.* (2020) can be dowloaded [here: `r emoji::emoji("point_right")` `r emoji::emoji("globe_with_meridians")` *NAR G&B* reproducible code files](https://github.com/borishejblum/dearseq/releases/download/v1.7.2/NARGAB_ReproducibleCodeFiles.zip)

> Agniel D & Hejblum BP (2017). Variance component score test for time-course gene set analysis of longitudinal RNA-seq data, [*Biostatistics*](https://academic.oup.com/biostatistics/article-abstract/18/4/589/3065599), **18**(*4*):589-604. [arXiv:1605.02351](https://arxiv.org/abs/1605.02351v4) [DOI: 10.1093/biostatistics/kxx005](https://doi.org/10.1093/biostatistics/kxx005)

More information about `dearseq` can be found [here: `r emoji::emoji("point_right")` `r emoji::emoji("globe_with_meridians")` pkgdown website](https://borishejblum.github.io/dearseq/)

## Installation

***`dearseq` is  available from [Bioconductor](https://www.bioconductor.org)***

```{r bioconductor-install, eval=FALSE}
#install.packages("BiocManager")
BiocManager::install("dearseq")
```

The development version of `dearseq` is available on  [GitHub](https://github.com/borishejblum/dearseq):
```{r GitHub-install, eval=FALSE}
#install.packages("BiocManager")
remotes::install_github("borishejblum/dearseq")
```

 -- Marine Gauthier, Denis Agniel & Boris Hejblum
